Effect of quality-adjusted production function inputs

### 2015-05-14 Articles/Presentation_ESEE_2015/ESEE_2015.Rnw
Cleaning up ESEE analysis using formulaUtils.

<<Data, eval=TRUE>>=
countries <- c("PT", "UK")
dataPT <- filter(IST, Country == "PT")
dataUK <- filter(Leeds, Country == "UK")
@

<<Time Series Variables>>=
# This block of code sets up the time-series variables 
# to be used for each country.
# Adding this level of abstraction makes it easier 
# for us to change variables later.
# It also makes it easier to see what is going on 
# in each formula throughout the document.
# Use of substitute establishes an R "name"
# that can be a component of a formula.

vars <-   
  list( 
    PT = list(
      response = list(unadjusted = "iGDP", adjusted = "iGDP"),
      capital = list(unadjusted ="iKstkS.L", adjusted = "iKservS.L"),
      labor = list(unadjusted = "iL", adjusted = "ihLest"),
      energy = list(unadjusted = "iXpMP", adjusted = "iUMP"),
      time = list(unadjusted = "iYear", adjusted = "iYear")
    ), 
    UK = list(
      response = list(unadjusted = "iGDP", adjusted = "iGDP"),
      capital = list(unadjusted = "iKstkO.WwithRD", adjusted = "iKservO.WwithRD"), 
      labor = list(unadjusted = "iL", adjusted = "ihLest"),
      energy = list(unadjusted = "iXp", adjusted = "iU"),
      time = list(unadjusted = "iYear", adjusted = "iYear")
    )
  )

# # Economic output
# y_PT <- substitute(iGDP)
# y_UK <- substitute(iGDP)
# # Portugal undajusted inputs
# k_PT_unadj <- substitute(iKstkS.L)
# l_PT_unadj <- substitute(iL)
# e_PT_unadj <- substitute(iXpMP)
# # Portugal quality-adjusted inputs
# k_PT_adj <- substitute(iKservS.L)
# l_PT_adj <- substitute(ihLest)
# e_PT_adj <- substitute(iUMP)
# # UK unadjusted inputs
# k_UK_unadj <- substitute(iKstkO.WwithRD)
# l_UK_unadj <- substitute(iL)
# e_UK_unadj <- substitute(iXp)
# # UK quality-adjusted inputs
# k_UK_adj <- substitute(iKservO.WwithRD)
# l_UK_adj <- substitute(ihLest)
# e_UK_adj <- substitute(iU)
# # Time variables
# t_PT <- substitute(iYear)
# t_UK <- substitute(iYear)

# vars <- list()
# # Economic output
# vars$y.PT <- substitute(iGDP)
# vars$y.UK <- substitute(iGDP)
# # Portugal undajusted inputs
# vars$k.PT.unadj <- substitute(iKstkS.L)
# vars$l.PT.unadj <- substitute(iL)
# vars$e.PT.unadj <- substitute(iXpMP)
# # Portugal quality-adjusted inputs
# vars$k.PT.adj <- substitute(iKservS.L)
# vars$l.PT.adj <- substitute(ihLest)
# vars$e.PT.adj <- substitute(iUMP)
# # UK unadjusted inputs
# vars$k.UK.unadj <- substitute(iKstkO.WwithRD)
# vars$l.UK.unadj <- substitute(iL)
# vars$e.UK.unadj <- substitute(iXp)
# # UK quality-adjusted inputs
# vars$k.UK.adj <- substitute(iKservO.WwithRD)
# vars$l.UK.adj <- substitute(ihLest)
# vars$e.UK.adj <- substitute(iU)
# # Time variables
# vars$t.PT <- substitute(iYear)
# vars$t.UK <- substitute(iYear)

# y <- list()
# k <- list()
# l <- list()
# e <- list()
# t <- list()
# # Economic output
# y$PT <- substitute(iGDP)
# y$UK <- substitute(iGDP)
# # Portugal undajusted inputs
# k$PT$unadj <- substitute(iKstkS.L)
# l$PT$unadj <- substitute(iL)
# e$PT$unadj <- substitute(iXpMP)
# # Portugal quality-adjusted inputs
# k$PT$adj <- substitute(iKservS.L)
# l$PT$adj <- substitute(ihLest)
# e$PT$adj <- substitute(iUMP)
# # UK unadjusted inputs
# k$UK$unadj <- substitute(iKstkO.WwithRD)
# l$UK$unadj <- substitute(iL)
# e$UK$unadj <- substitute(iXp)
# # UK quality-adjusted inputs
# k$UK$adj <- substitute(iKservO.WwithRD)
# l$UK$adj <- substitute(ihLest)
# e$UK$adj <- substitute(iU)
# # Time variables
# t$PT <- substitute(iYear)
# t$UK <- substitute(iYear)
# 
# foo <- function(country = c("UK", "PT"), adjusted=TRUE, energy=TRUE, envir=parent.frame()) {
#   country <- match.arg(country)
#   type <- if (adjusted) "adj" else "unadj"
#   res <- as.formula(
#       paste(y[[country]], 
#             paste(k[[country]][[type]], l[[country]][[type]], if (energy) e[[country]][[type]] else NULL, t[[country]], sep="+"), 
#             sep="~")
#     )
#   environment(res) <- envir
#   res
# }
# 
# 
# for (country in c("PT", "UK")) {
#   for (adjusted in c(TRUE, FALSE)) {
#     for (energy in c(TRUE, FALSE)) {
#       print(foo(country = country, adjusted = adjusted, energy=energy))
#     }
#   }
# }
#

# 
# countries <- c(substitute(PT), substitute(UK))
# adjusteds <- c(substitute(unadj), substitute(adj))
# energies <- c(substitute(noE), substitute(withE))
# f <- list()
# for (country in c(substitute(PT), substitute(UK))){
#   for (adjusted in adjusteds){
#     for (energy in energies){
#       if (energy == substitute(noE)){
#         eval(substitute(f$ctry$adjst$ener <- y$ctry ~ k$ctry$adjst + l$ctry$adjst + t$ctry, 
#                         list(ctry=country, adjst=adjusted, ener=energy)))
#       } else if (energy == substitute(withE)){
#         eval(substitute(f$ctry$adjst$ener <- y$ctry ~ k$ctry$adjst + l$ctry$adjst + e$ctry$adjst + t$ctry, 
#                         list(ctry=country, adjst=adjusted, ener=energy)))
#       } else {
#         stop(paste("Unknown energy", energy, "in loop for f"))
#       }
#     }
#   }
# }

@@

dataPT <- transform(dataPT, lPerGDP.unadj = eval(k_PT_unadj)/iGDP)
dataPT <- transform(dataPT, lPerGDP.unadj = eval(l_PT_unadj)/iGDP)
dataPT <- transform(dataPT, ePerGDP.unadj = eval(e_PT_unadj)/iGDP)
dataUK <- transform(dataUK, kPerGDP.unadj = eval(k_UK_unadj)/iGDP)
dataUK <- transform(dataUK, lPerGDP.unadj = eval(l_UK_unadj)/iGDP)
dataUK <- transform(dataUK, ePerGDP.unadj = eval(e_UK_unadj)/iGDP)
# Quality-adjusted variables
dataPT <- transform(dataPT, kPerGDP.adj = eval(k_PT_adj)/iGDP)
dataPT <- transform(dataPT, lPerGDP.adj = eval(l_PT_adj)/iGDP)
dataPT <- transform(dataPT, ePerGDP.adj = eval(e_PT_adj)/iGDP)
dataUK <- transform(dataUK, kPerGDP.adj = eval(k_UK_adj)/iGDP)
dataUK <- transform(dataUK, lPerGDP.adj = eval(l_UK_adj)/iGDP)
dataUK <- transform(dataUK, ePerGDP.adj = eval(e_UK_adj)/iGDP)

@@

<<Formulas>>=
# Set up the formulas for each type of analysis.
# f <- NULL
# f$PT$unadj$noE <- as.formula(
#   substitute( y ~ k + l + t,
#               list( y = y_PT, k = k_PT_unadj, 
#                     l = l_PT_unadj, t = t_PT)))
# f$PT$unadj$withE <- as.formula(
#   substitute( y ~ k + l + e + t,
#               list( y = y_PT, k = k_PT_unadj,
#                     l = l_PT_unadj, e = e_PT_unadj, t = t_PT)))
# f$PT$adj$noE <- as.formula(
#   substitute( y ~ k + l + t,
#               list( y = y_PT, k = k_PT_adj,
#                     l = l_PT_adj, t = t_PT)))
# f$PT$adj$withE <- as.formula(
#   substitute( y ~ k + l + e + t,
#               list( y = y_PT, k = k_PT_adj,
#                     l = l_PT_adj, e = e_PT_adj, t = t_PT)))
# f$UK$unadj$noE <- as.formula(
#   substitute( y ~ k + l + t,
#               list( y = y_UK, k = k_UK_unadj,
#                     l = l_UK_unadj, t = t_UK)))
# f$UK$unadj$withE <- as.formula(
#   substitute( y ~ k + l + e + t,
#               list( y = y_UK, k = k_UK_unadj,
#                     l = l_UK_unadj, e = e_UK_unadj, t = t_UK)))
# f$UK$adj$noE <- as.formula(
#   substitute( y ~ k + l + t,
#               list( y = y_UK, k = k_UK_adj,
#                     l = l_UK_adj, t = t_UK)))
# f$UK$adj$withE <- as.formula(
#   substitute( y ~ k + l + e + t,
#               list( y = y_UK, k = k_UK_adj,
#                     l = l_UK_adj, e = e_UK_adj, t = t_UK)))

makeFormula <- function(time.series, country, adjusted, energy, envir=parent.frame()) {
  country <- match.arg(country, names(time.series))
  if (! adjusted %in% c(FALSE, TRUE)){
    stop("adjusted must be a boolean")
  }
  if (! energy %in% c(FALSE, TRUE)){
    stop("energy must be a boolean")
  }
  type <- if (adjusted) "adjusted" else "unadjusted"
  res <- as.formula( 
    gsub( "\\+ *\\+", "+ ", 
          paste(time.series[[country]][["response"]][[type]], 
                paste(time.series[[country]][["capital"]][[type]], 
                      time.series[[country]][["labor"]][[type]], 
                      if (energy) {
                        time.series[[country]][["energy"]][[type]]
                      } else {
                        NULL
                      }, 
                      time.series[[country]][["time"]][[type]], sep="+"), 
                sep="~")
    )
  )
  environment(res) <- envir
  return(res)
}

f <- list()
for (country in countries) {
  for (adjusted in c(FALSE, TRUE)) {
    for (energy in c(FALSE, TRUE)) {
      newf <- makeFormula(vars, country = country, adjusted = adjusted, energy=energy)
      namef <- paste(country, if(adjusted) "adj" else "unadj", if(energy) "withE" else "noE", sep=".")
      newnames <- c(names(f), namef)
      f <- c(f, newf)
      names(f) <- newnames
    }
  }
}
@

@@

\begin{table} \caption{Unadjusted time-series variables.} 
\label{tab:unadjusted_variables} 
  \begin{center}
    \begin{tabular}{r l l} 
      \toprule
      Variable    & Portugal           & UK                 \\
      \midrule
      Output      & \Sexpr{vars$PT$response$unadjusted}       & \Sexpr{vars$UK$response$unadjusted}       \\
      Capital     & \Sexpr{vars$PT$capital$unadjusted} & \Sexpr{vars$UK$capital$unadjusted} \\
      Labor       & \Sexpr{vars$PT$labor$unadjusted} & \Sexpr{vars$UK$labor$unadjusted} \\
      Energy      & \Sexpr{vars$PT$energy$unadjusted} & \Sexpr{vars$UK$energy$unadjusted} \\
      \bottomrule
    \end{tabular}
  \end{center}
\end{table}

@@

\begin{table} \caption{Quality-adjusted time series variables.} 
\label{tab:quality_adjusted_variables} 
  \begin{center}
    \begin{tabular}{r l l} 
      \toprule
      Variable    & Portugal         & UK               \\
      \midrule
      Output      & \Sexpr{vars$PT$response$adjusted}       & \Sexpr{vars$UK$response$adjusted}       \\
      Capital     & \Sexpr{vars$PT$capital$adjusted} & \Sexpr{vars$UK$capital$adjusted} \\
      Labor       & \Sexpr{vars$PT$labor$adjusted} & \Sexpr{vars$UK$labor$adjusted} \\
      Energy      & \Sexpr{vars$PT$energy$adjusted} & \Sexpr{vars$UK$energy$adjusted} \\
      \bottomrule
    \end{tabular}
  \end{center}
\end{table}

@@

<<Historical Data, fig.pos="H", fig.align='center', fig.width=6.5, fig.height=4, fig.cap="Historical data.", warning=FALSE>>=
unadjVars_PT <- c(vars$PT$response$unadjusted, 
                  vars$PT$capital$unadjusted, 
                  vars$PT$labor$unadjusted, 
                  vars$PT$energy$unadjusted)
unadjVars_UK <- c(vars$UK$response$unadjusted, 
                  vars$UK$capital$unadjusted, 
                  vars$UK$labor$unadjusted, 
                  vars$UK$energy$unadjusted)
commonVars <- c("Year", "iYear", "Country")
unadjData_PT <- historicalData[ , unique(c(unadjVars_PT, commonVars))] %>% filter(Country=="PT")
unadjData_UK <- historicalData[ , unique(c(unadjVars_UK, commonVars))] %>% filter(Country=="UK")
unadjData <- plyr::rbind.fill(unadjData_PT, unadjData_UK)
unadjData$type <- "Unadjusted"
adjVars_PT <- c(vars$PT$response$adjusted, 
                vars$PT$capital$adjusted, 
                vars$PT$labor$adjusted, 
                vars$PT$energy$adjusted)
adjVars_UK <- c(vars$UK$response$adjusted, 
                vars$UK$capital$adjusted, 
                vars$UK$labor$adjusted, 
                vars$UK$energy$adjusted)
adjData_PT <- historicalData[ , unique(c(adjVars_PT, commonVars))] %>% filter(Country=="PT")
adjData_UK <- historicalData[ , unique(c(adjVars_UK, commonVars))] %>% filter(Country=="UK")
adjData <- plyr::rbind.fill(adjData_PT, adjData_UK)
adjData$type <- "Quality-adjusted"

trimmedData <- plyr::rbind.fill(unadjData, adjData)

@@

# Relevel factors before plotting
mdata$Country <- relevelFactor(as.factor(mdata$Country), c("PT", "UK"))
mdata$type <- relevelFactor(as.factor(mdata$type), c("Unadjusted", "Quality-adjusted"))
mdata$variable <- relevelFactor(as.factor(mdata$variable), unique(
  c(vars$PT$response$unadjusted, vars$PT$capital$unadjusted, vars$PT$labor$unadjusted, vars$PT$energy$unadjusted, 
    vars$PT$response$adjusted,   vars$PT$capital$adjusted,   vars$PT$labor$adjusted,   vars$PT$energy$adjusted, 
    vars$UK$response$unadjusted, vars$UK$capital$unadjusted, vars$UK$labor$unadjusted, vars$UK$energy$unadjusted, 
    vars$UK$response$adjusted,   vars$UK$capital$adjusted,   vars$UK$labor$adjusted,   vars$UK$energy$adjusted)))
historicalPlot(mdata, 
               mapping=aes(x=Year, y=value, group=variable, linetype=variable), 
               facet_formula = Country ~ type, 
               line_types = c(1,2,4,3,2,4,3,2,3,2,3)) + 
  scale_y_continuous(breaks=c(1, 2, 4, 6, 8, 10, 12)) +
  coord_cartesian(ylim=c(0.5, 13)) +
  xlab("") + 
  ylab("Indexed (1960=1)")
@

>>>>

<<Data>>=
countries <- c("PT", "UK")
dataPT <- filter(IST, Country == "PT")
dataUK <- filter(Leeds, Country == "UK")

@@

dataPT <- within(filter(IST, Country == "PT"), {
  kPerGDP.unadj <- iKstkS.L / iGDP # Works
#   kPerGDP.unadj <- as.name(extract_rhs(forms$PT$unadjusted, 1)) / as.name(extract_lhs(forms$PT$unadjusted, 1)) # Fails
})

dataPT <- transform(dataPT, kPerGDP.unadj = eval(k_PT_unadj)/iGDP)
dataPT <- transform(dataPT, lPerGDP.unadj = eval(l_PT_unadj)/iGDP)
dataPT <- transform(dataPT, ePerGDP.unadj = eval(e_PT_unadj)/iGDP)
dataUK <- transform(dataUK, kPerGDP.unadj = eval(k_UK_unadj)/iGDP)
dataUK <- transform(dataUK, lPerGDP.unadj = eval(l_UK_unadj)/iGDP)
dataUK <- transform(dataUK, ePerGDP.unadj = eval(e_UK_unadj)/iGDP)
# Quality-adjusted variables
dataPT <- transform(dataPT, kPerGDP.adj = eval(k_PT_adj)/iGDP)
dataPT <- transform(dataPT, lPerGDP.adj = eval(l_PT_adj)/iGDP)
dataPT <- transform(dataPT, ePerGDP.adj = eval(e_PT_adj)/iGDP)
dataUK <- transform(dataUK, kPerGDP.adj = eval(k_UK_adj)/iGDP)
dataUK <- transform(dataUK, lPerGDP.adj = eval(l_UK_adj)/iGDP)
dataUK <- transform(dataUK, ePerGDP.adj = eval(e_UK_adj)/iGDP)

@@

<<Formulas>>=
# Set up the formulas for each type of analysis.
forms <- 
  list( 
    PT = c(unadjusted = build_formula("iGDP", list("iKstkS.L", "iL", "iXpMP", "iYear")),
             adjusted = build_formula("iGDP", list("iKservS.L", "ihLest", "iUMP", "iYear"))),
    UK = c(unadjusted = build_formula("iGDP", list("iKservO.WwithRD", "iL", "iXp", "iYear")),
             adjusted = build_formula("iGDP", list("iKservO.WwithRD", "ihLest", "iU", "iYear")))
  )

@

@@

\begin{table} \caption{Unadjusted time-series variables.} 
\label{tab:unadjusted_variables} 
  \begin{center}
    \begin{tabular}{r l l} 
      \toprule
      Variable    & Portugal           & UK                 \\
      \midrule
      Output      & \Sexpr{deparse(lhs(forms$PT$unadjusted))}   & \Sexpr{deparse(lhs(forms$UK$unadjusted))}   \\
      Capital     & \Sexpr{extract_rhs(forms$PT$unadjusted, 1)} & \Sexpr{extract_rhs(forms$UK$unadjusted, 1)} \\
      Labor       & \Sexpr{extract_rhs(forms$PT$unadjusted, 2)} & \Sexpr{extract_rhs(forms$UK$unadjusted, 2)} \\
      Energy      & \Sexpr{extract_rhs(forms$PT$unadjusted, 3)} & \Sexpr{extract_rhs(forms$UK$unadjusted, 3)} \\
      \bottomrule
    \end{tabular}
  \end{center}
\end{table}

@@

\begin{table} \caption{Quality-adjusted time series variables.} 
\label{tab:quality_adjusted_variables} 
  \begin{center}
    \begin{tabular}{r l l} 
      \toprule
      Variable    & Portugal         & UK               \\
      \midrule
      Output      & \Sexpr{deparse(lhs(forms$PT$adjusted))}   & \Sexpr{deparse(lhs(forms$UK$adjusted))}   \\
      Capital     & \Sexpr{extract_rhs(forms$PT$adjusted, 1)} & \Sexpr{extract_rhs(forms$UK$adjusted, 1)} \\
      Labor       & \Sexpr{extract_rhs(forms$PT$adjusted, 2)} & \Sexpr{extract_rhs(forms$UK$adjusted, 2)} \\
      Energy      & \Sexpr{extract_rhs(forms$PT$adjusted, 3)} & \Sexpr{extract_rhs(forms$UK$adjusted, 3)} \\
      \bottomrule
    \end{tabular}
  \end{center}
\end{table}

@@

<<Historical Data, fig.pos="H", fig.align='center', fig.width=6.5, fig.height=4, fig.cap="Historical data.", warning=FALSE>>=
unadjVars_PT <- all.vars(forms$PT$unadjusted)[1:4]
unadjVars_UK <- all.vars(forms$UK$unadjusted)[1:4]
commonVars <- c("Year", "iYear", "Country")
unadjData_PT <- historicalData[ , unique(c(unadjVars_PT, commonVars))] %>% filter(Country=="PT")
unadjData_UK <- historicalData[ , unique(c(unadjVars_UK, commonVars))] %>% filter(Country=="UK")
unadjData <- plyr::rbind.fill(unadjData_PT, unadjData_UK)
unadjData$type <- "Unadjusted"
adjVars_PT <- all.vars(forms$PT$adjusted)[1:4]
adjVars_UK <- all.vars(forms$UK$adjusted)[1:4]
adjData_PT <- historicalData[ , unique(c(adjVars_PT, commonVars))] %>% filter(Country=="PT")
adjData_UK <- historicalData[ , unique(c(adjVars_UK, commonVars))] %>% filter(Country=="UK")
adjData <- plyr::rbind.fill(adjData_PT, adjData_UK)
adjData$type <- "Quality-adjusted"
trimmedData <- plyr::rbind.fill(unadjData, adjData)

@@

# Relevel factors before plotting
mdata$Country <- relevelFactor(as.factor(mdata$Country), c("PT", "UK"))
mdata$type <- relevelFactor(as.factor(mdata$type), c("Unadjusted", "Quality-adjusted"))
mdata$variable <- relevelFactor(as.factor(mdata$variable), unique(c(unadjVars_PT, adjVars_PT, unadjVars_UK, adjVars_UK)))

historicalPlot(mdata, 
               mapping=aes(x=Year, y=value, group=variable, linetype=variable), 
               facet_formula = Country ~ type, 
               line_types = c(1,2,4,3,2,4,3,2,3,2,3)) + 
  scale_y_continuous(breaks=c(1, 2, 4, 6, 8, 10, 12)) +
  coord_cartesian(ylim=c(0.5, 13)) +
  xlab("") + 
  ylab("Indexed (1960=1)")
@
